---
title: HiRise Observations of Mars
author: "Micah J Waldstein"
output:
  html_document:
    fig_width: 8.5
    fig_align: "center"
    fig_show: "hold"
    cache: TRUE
    toc: FALSE
    self_contained: FALSE
    results: "hold"
---
```{r set_seed, echo = F}
# Ensure stable output
set.seed(7035)
```
*DISCLAIMER: While I know a thing or two, there's a reasonable chance I got
some things wrong or at very least there are certainly more efficient ways to
go about things. Feedback always appreciated!*

After
[two](https://micah.waldste.in/blog/2018/06/introduction-to-spherical-densities-in-r/)
[posts]()
on how to analyze the density of points on a globe, we get to what I was really
curious about - the density of images of Mars made by
[HiRISE](https://hirise.lpl.arizona.edu/) on the
[Mars Reconnaissance Orbiter](https://en.wikipedia.org/wiki/Mars_Reconnaissance_Orbiter).

All the records of the (still ongoing) mission are available publicly from the
[Planetary Data System](https://pds.jpl.nasa.gov/index.shtml), specifically
[here](https://hirise-pds.lpl.arizona.edu/PDS/).

I'll be using a few datasets in this analysis:
 * [RDRCUMINDEX.LBL](https://hirise-pds.lpl.arizona.edu/PDS/INDEX/RDRCUMINDEX.LBL) -
   describes the fields of the Reduced Data Records (RDR), where images have
   been corrected and geometrically mapped. In pds3 format.
 * [RDRCUMINDEX.TAB](https://hirise-pds.lpl.arizona.edu/PDS/INDEX/RDRCUMINDEX.TAB) -
   A csv with all the observations.

We'll be using all the tricks I covered in my posts on spherical densities and
the [pds3 library](https://github.com/mwaldstein/pds3) I created and maintain.

# Set-Up

First, we'll make use of a number of libraries and setup our plotting
environment and the pds3 reader:
```{r libraries, message = F}
library(rgdal)       # For coordinate transforms
library(sp)          # For plotting grid images
library(sf)
library(lwgeom)
library(Directional) # For spherical density functions
library(raster)
library(pds3)
```

## Observation Data

First, we'll read in the observations and the column labels. We'll also pull
out the positions so we have a simple dataset ready to go. Finally we'll create
a `SpatialPoints` object for plotting.

```{r observation_data, cache = TRUE}
  header_label <- pds3_read(file.path("..", "data", "RDRCUMINDEX.LBL"))
  headers <- do.call(rbind, header_label$odl$RDR_INDEX_TABLE$COLUMN)
  head(headers)

  observations <- read.table(file.path("..", "data", "RDRCUMINDEX.TAB"),
                             sep = ",")
  names(observations) <- unlist(headers[,1])
  head(observations)

  positions <- observations[,c("CORNER1_LONGITUDE", "CORNER1_LATITUDE")]
  names(positions) <- c("lon", "lat")
  positions$lon[positions$lon < 0] <- 360 + positions$lon[positions$lon < 0]
  positions$lon <- positions$lon - 180

  sd_points <- SpatialPoints(positions,
                             proj4string = CRS("+proj=longlat +datum=WGS84 +no_defs"))
```

## Calculate Densities

```{r basic_plot}
grid.size <- 60
r.template <- raster(extent(sd_points), nrow = grid.size, ncol = grid.size,
                     crs = "+proj=longlat +datum=WGS84")
r <- rasterize(sd_points, r.template, field = 1, fun = "count")

plot(r)
plot(gridlines(r), add = TRUE, col = "grey30", alpha = .1)
contour(r, add = TRUE)
```

## Using Polygons

We'll use raster data again, but we'll immediately convert it into a grid of
square polygons which we can then project

```{r obs_projected}
r2 <- r
# We'll manually colorize
r2 <- cut(r2,
          pretty(r2[], 50),
          include.lowest = F)
color.vals <- rev(terrain.colors(50))
pol <- rasterToPolygons(r2)
crs1 <- "+proj=wintri"
pol.crs1 <- spTransform(pol, crs1)
plot(pol.crs1, col=color.vals[r2[]], border = NA)
plot(gridlines(pol.crs1), add = TRUE, col = "grey30", alpha = .1)
```

Now that looks good!

One thing to keep in mind however - because our polygons are rectangular in equal
coordinates, they will warp and distort as a projection gets more severe. In
our animation, you can see what I mean

## Animating

We're projecting into an orthographic projection to simulate the rotating
globe. A few things you'll see in the code where I jump through hoops:
 * **Cropping the top** - If I leave the top polygons in place, they bunch up
   in an ugly fashion
 * **Making features valid** - Both for the world and our heatmap polygons I
   jump through hoops to make sure only valid polygons get through to the final
   plot.


```{r earthquake_ani, fig.show='animate', ffmpeg.format='gif', fig.width = 4.5, interval = .2, warning=F, eval = F}
r3 <- raster(sGDF)

# Crop down because projecting the poles causes problems
r.crop <- res(r3)
rc <- crop(r3, extent(-180, 180,
                      -90 + r.crop[2], 90 - r.crop[2]))
pol <- rasterToPolygons(rc)
pol.breaks <- pretty(pol$att, 20)
pol.colors <- rev(terrain.colors(length(pol.breaks) - 1))
# Make the lowest color transparent
substr(pol.colors[1], 8, 9) <- "00"

par_old <- par()
par(mar = c(0, 0, 0, 0))
n.frames <- 30
grad <- st_graticule(ndiscr = 1e4)
for (i in 1:n.frames) {
  long <- -180 + (i - 1) * 360 / n.frames
  crs.ani <- paste0("+proj=ortho +lat_0=0 +lon_0=", long)
  grad.ani <- st_geometry(st_transform(grad, crs.ani))

  world.ani <- st_transform(st_geometry(world), crs = crs.ani)
  world.ani <- st_make_valid(world.ani)
  # We don't want the points
  world.ani <- world.ani[st_geometry_type(world.ani) %in% c('POLYGON',
                                                            'MULTIPOLYGON')]

  # There are inevitable some bad polygons out of the transform
  world.ani <- world.ani[st_is_valid(world.ani)]

  pol.ani <- st_transform(as(pol, "sf"), crs.ani)
  pol.ani.geo <- lwgeom::st_make_valid(pol.ani)
  pol.ani.geo <- pol.ani.geo[st_geometry_type(pol.ani.geo) %in% c('POLYGON',
                                                                  'MULTIPOLYGON',
                                                                  'GEOMETRYCOLLECTION'), ]
  pol.ani.geo <- pol.ani.geo[st_is_valid(pol.ani.geo), ]
  pol.ani.geo <- pol.ani.geo[!st_is_empty(pol.ani.geo), ]

  plot(grad.ani, col = "black")
  plot(world.ani, add = TRUE, col = "grey30", border = "grey")
  plot(pol.ani.geo, border = NA, breaks = pol.breaks, pal = pol.colors,
       add = TRUE, main = NA, key.pos = NULL)
}
par(par_old)
```

Looks pretty good, but we do have some interesting world map problems with
countries popping out as they reach the edge... Something to investigate
another day.

## Final Notes

In both these examples we've used global data as it shows the problems of using
"traditional" density estimators, but the same issue exists at all scales. It
is just a question of when a simpler approximation is reasonable.

You can also see a bit of blockiness which we could reduce with an increase in
grid size, but that will be very dependent on need.

Next, some real data...

# Appendix

## Spherical Density Function

This calculates a grid of densities which can then be used with `geom_contour`.
The code basically comes directly from [Directional's
vmf.kerncontour](https://rdrr.io/cran/Directional/man/vmf.kerncontour.html),
only returning a data.frame instead of actually plotting the output.

```{r vmf_kerncontour}
vmf.kerncontour.new <- function(u, thumb = "none", ret.all = FALSE, full = FALSE,
                            ngrid = 100) {
  ## u contains the data in latitude and longitude
  ## the first column is the latitude and the
  ## second column is the longitude
  ## thumb is either 'none' (default), or 'rot' (Garcia-Portugues, 2013)
  ## ret.all if set to TRUE returns a matrix with latitude, longitude and density
  ## full if set to TRUE calculates densities for the full sphere, otherwise
  ##   using extents of the data
  ## ngrid specifies the number of points taken at each axis
  n <- dim(u)[1]  ## sample size
  x <- euclid(u)

  if (thumb == "none") {
    h <- as.numeric( vmfkde.tune(x, low = 0.1, up = 1)[1] )
  } else if (thumb == "rot") {
    k <- vmf(x)$kappa
    h <- ( (8 * sinh(k)^2) / (k * n * ( (1 + 4 * k^2) * sinh(2 * k) -
    2 * k * cosh(2 * k)) ) ) ^ ( 1/6 )
  }

  if (full) {
    x1 <- seq( 0, 180, length = ngrid )  ## latitude
    x2 <- seq( 0, 360, length = ngrid )  ## longitude
  } else {
    x1 <- seq( min(u[, 1]) - 5, max(u[, 1]) + 5, length = ngrid )  ## latitude
    x2 <- seq( min(u[, 2]) - 5, max(u[, 2]) + 5, length = ngrid )  ## longitude
  }
  cpk <- 1 / (  ( h^2)^0.5 *(2 * pi)^1.5 * besselI(1/h^2, 0.5) )
  mat <- matrix(nrow = ngrid, ncol = ngrid)

  for (i in 1:ngrid) {
    for (j in 1:ngrid) {
      y <- euclid( c(x1[i], x2[j]) )
      a <- as.vector( tcrossprod(x, y / h^2) )
      can <- sum( exp(a + log(cpk)) ) / ngrid
      if (abs(can) < Inf)   mat[i, j] <- can
    }
  }

  if (ret.all) {
    return(list(Lat = x1, Long = x2, h = h, d = mat))
  } else {
    contour(mat$Lat, mat$Long, mat, nlevels = 10, col = 2, xlab = "Latitude",
            ylab = "Longitude")
    points(u[, 1], u[, 2])
  }
}
```

## References

 * McEwen, A., Mars Reconnaissance Orbiter High Resolution Imaging Science
   Experiment, Reduced Data Record, MRO-M-HIRISE-3-RDR-V1.0, NASA Planetary Data
   System, 2007.
